Finiteness in Cubical Type Theory

### 2020-06-02 sections/finiteness-predicates.tex
Update finiteness-predicates.tex

\begin{agdalisting} \label{bool-slop}
  Consider the following two proofs of the finiteness of Bool:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-slop}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  While they both represent the ``same'' information (i.e. they both tell us
  that bool is finite), they are nonetheless not definitionally equal.
\end{agdalisting}

@@

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

To remedy the problem, we will use propositional truncation
(def.~\ref{prop-trunc}).
\begin{definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite if there exists a propositionally truncated
  proof that \(A\) is manifest Bishop finite or equivalent to a finite prefix of
  the natural numbers.
  \begin{equation}
    \mathcal{C}(A) \coloneqq \lVert \mathcal{B}(A) \rVert \simeq \lVert \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq A) \rVert
  \end{equation}
\end{definition}
At first glance, it might seem that we lose any useful properties we could
derive from \(\mathcal{B}\).
Luckily, this is not the case: by eliminator \ref{elim-prop-coh} of
def.~\ref{prop-trunc}, we need only show that the output is uniquely determined.

The following two lemmas are proven in
\cite{yorgeyCombinatorialSpeciesLabelled2014} (Proposition 2.4.9 and 2.4.10,
respectively), in much the same way as we have done here.
Our contribution for this section is simply the formalisation.
\begin{lemma}
  Given a cardinally finite type, we can derive the type's cardinality, as well
  as a propositionally truncated proof of equivalence with \(\textbf{Fin}\)s of
  the same cardinality.
  \begin{equation}
    \mathcal{C}(A) \rightarrow \Sigma {(n : \mathbb{N})} , \lVert \textbf{Fin}(n) \simeq A \rVert
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a cardinally-finite type, with proof \(F : \mathcal{C}(A)\).
  Our task is to extract a natural number \(n : \mathbb{N}\) representing the
  cardinality of \(A\), and a propositionally-truncated proof that \(A\) is
  equivalent to \(\mathbf{Fin}(n)\).

@@

  This is a well-known chestnut in dependently-typed programming, and one that
  has a surprisingly tricky and complex proof.
  We do not include it here, since it has already been explored elsewhere, but
  it is present in our formalisation.
\end{proof}
\begin{lemma} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{lemma}
\begin{proof}
  Since we can already derive decidable equality from a proof of manifest Bishop
  finiteness, it suffices to show that decidable equality is itself a
  proposition.
  \begin{equation}
    \text{isProp}(\Pi(x, y : A) , \mathbf{Dec}(x \equiv y))
  \end{equation}
  First, it is clear that \(x \equiv y\) is a proposition: since the type \(A\)
  has decidable equality, by Hedburg's theorem it is a set,
  meaning precisely that \(x \equiv y\) is a proposition.

@@

\begin{definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) \coloneqq \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{definition}
As with manifest Bishop finiteness, the only difference with this type and split
enumerability is the membership proof: here we have propositionally truncated
it.
This has two effects.
First, it means that this proof represents a true surjection (rather than a
split surjection) from \(\mathbf{Fin}\).
\begin{lemma}
  \begin{equation}
    \mathcal{E}(A) \simeq \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \twoheadrightarrow A)
  \end{equation}
\end{lemma}
The proof for this lemma is similar in structure to
lemma~\ref{split-enum-is-split-surj} and lemma~\ref{bishop-equiv}.

Secondly, it means the predicate does not imply decidable equality.
More significantly, it allows the predicate to be defined over non-set types,
like the circle.
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{equation}
    \begin{alignedat}{3}
      S^1 \coloneqq & \; \text{base} &&: S^1 ; \\
      | & \; \text{loop} &&: \text{base} \equiv \text{base} ; 
    \end{alignedat}
  \end{equation}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{definition}
\begin{lemma}
  The circle \(S^1\) is manifestly enumerable.
\end{lemma}
\subsection{Relation To Split Enumerability }
\begin{lemma} \label{split-enum-to-manifest-enum}
  Any split enumerable type is manifestly enumerable.
\end{lemma}
This lemma is proven by truncating the membership proof in split enumerability.

\begin{lemma} \label{manifest-enum-to-split-enum}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{lemma}
\section{Kuratowski Finiteness}
The first thing we must define is a representation of subsets.
\begin{definition}[Kuratowski Finite Subset]
  \(\mathcal{K}(A)\) is the type of Kuratowski-finite subsets of \(A\).
  \begin{inductivetype}{\mathcal{K}(A)}
    \inductivetypeclause{\left[ \right]}{\mathcal{K}(A)}
    \inductivetypeclause{\wc \dblcolon \wc}{A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A)}
    \inductivetypeclause{\text{com}}{\Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{dup}}{\Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{trunc}}{\Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q}
  \end{inductivetype}
  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

>>>>

\begin{agdalisting} \label{bool-slop}
  Consider the following proof of the finiteness of Bool:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-slop}
  While it represents the ``same'' information as our previous proof, it is
  clearly not the same \emph{object}.
\end{agdalisting}

@@

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst} \smallskip
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

To remedy the problem, we will use propositional truncation
(def.~\ref{prop-trunc}).
\begin{definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite if there exists a propositionally truncated
  proof that \(A\) is manifest Bishop finite or equivalent to a finite prefix of
  the natural numbers.
  \begin{equation}
    \mathcal{C}(A) \coloneqq \lVert \mathcal{B}(A) \rVert \simeq \lVert \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq A) \rVert
  \end{equation}
\end{definition}
At first glance, it might seem that we lose any useful properties we could
derive from \(\mathcal{B}\).
Luckily, this is not the case: by eliminator \ref{elim-prop-coh} of
def.~\ref{prop-trunc}, we need only show that the output is uniquely determined.
\subsection{Deriving Uniquely-Determined Quantities}
The following two lemmas are proven in
\cite{yorgeyCombinatorialSpeciesLabelled2014} (Proposition 2.4.9 and 2.4.10,
respectively), in much the same way as we have done here.
Our contribution for this section is simply the formalisation.
\begin{lemma}
  Given a cardinally finite type, we can derive the type's cardinality, as well
  as a propositionally truncated proof of equivalence with \(\textbf{Fin}\)s of
  the same cardinality.
  \begin{equation}
    \mathcal{C}(A) \rightarrow \Sigma {(n : \mathbb{N})} , \lVert \textbf{Fin}(n) \simeq A \rVert
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a cardinally-finite type, with proof \(F : \mathcal{C}(A)\).
  Our task is to extract a natural number \(n : \mathbb{N}\) representing the
  cardinality of \(A\), and a propositionally-truncated proof that \(A\) is
  equivalent to \(\mathbf{Fin}(n)\).

@@

  This is a well-known chestnut in dependently-typed programming, and one that
  has a surprisingly tricky and complex proof.
  We do not include it here, since it has already been explored elsewhere, but
  it is present in our formalisation.
\end{proof}

In order to prove that cardinal finiteness implies decidable equality, we will
need to show that decidable equality itself is a proposition.
In doing that we will use the following lemma:
\begin{lemma} \label{prop-refute}
  We can ``refute'' a propositionally-truncated proof of some proposition with a
  proof that the non-truncated proposition is false.
  \begin{equation}
    \frac{\neg A \; \; \; \lVert A \rVert}{\bot}
  \end{equation}
\end{lemma}
\begin{proof}
  We know we can eliminate from any value of type \(\lVert A \rVert\) into some
  \(B\) with a function \(A \to B\) if \(B\) is a proposition.
  That's precisely what we do in this case: \(\neg A\) is a function of type \(A
  \to \bot\), and we know that \(\bot\) is a proposition.
\end{proof}

\begin{lemma} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{lemma}
\begin{proof}
  Since we can already derive decidable equality from a proof of manifest Bishop
  finiteness, it suffices to show that decidable equality is itself a
  proposition.
  \begin{equation}
    \text{isProp}(\Pi(x, y : A) , \mathbf{Dec}(x \equiv y))
  \end{equation}
  First, it is clear that \(x \equiv y\) is a proposition: since the type \(A\)
  has decidable equality, by Hedburg's theorem it is a set,
  meaning precisely that \(x \equiv y\) is a proposition.

@@

\begin{definition}[Manifest Enumerability]
  Manifest enumerability is an enumeration predicate like Bishop finiteness or
  split enumerability with the only difference being a propositionally truncated
  membership proof.
  \begin{equation}
    \mathcal{E}(A) \coloneqq \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{definition}
As a function-based definition, this predicate represents surjections.
\begin{lemma}
  Manifest enumerability is equivalent to a surjection from a finite prefix of
  the natural numbers.
  \begin{equation}
    \mathcal{E}(A) \simeq \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \twoheadrightarrow A)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
    && \text{def.~\ref{split-enum-def} }(\mathcal{E})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \lVert \text{fib}_{\text{snd}(\mathit{xs})}(x) \rVert
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{surj-eqn} (surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \twoheadrightarrow A )
    && \text{eqn.~\ref{surj-arrow-eqn} } (\twoheadrightarrow) \; \qedhere
  \end{align*}
\end{proof}
\subsection{Instances for Non-Set Types}
The truncation has another very important implication: it means that the
predicate doesn't provide decidable equality on the underlying type.
Remember, this is how we knew that our previous predicates wouldn't allow for
non-set types: because they implied decidable equality, they also implied that
all conforming types had homotopy levels of at most 2. \todo{Are we doing the
  homotopy levels starting from -2 thing or from 0?}
This suggests that non-set types like the circle could conform to this
finiteness predicate.
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{equation}
    \begin{alignedat}{3}
      S^1 \coloneqq & \; \text{base} &&: S^1 ; \\
      | & \; \text{loop} &&: \text{base} \equiv \text{base} ; 
    \end{alignedat}
  \end{equation}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{definition}
\begin{lemma}
  The circle \(S^1\) is manifestly enumerable.
\end{lemma}
\begin{proof}
  The support list firstly is a list containing the point constructor for the
  circle.
  Since the cover proof is truncated, we need only consider the point
  constructors of the circle: as such, the cover proof is essentially the same
  as the one for \(\mathcal{E}!(\top)\).
\end{proof}
\subsection{Relation To Split Enumerability}
It is trivially easy to construct a proof that any split enumerable type is
manifest enumerable: we simply truncate the membership proof.
Going the other was requires us to extract a non-truncated proof from a
truncated one.
This proof relies on the following lemma:
\begin{lemma}
  We can ``recompute'' a truncated proof given a decision over a proof of the
  same type.
  \begin{equation}
    \frac{\lVert A \rVert \; \; \; \mathbf{Dec}(A)}{A}
  \end{equation}
\end{lemma}
\begin{proof}
  We proceed by case-analysis over the decision over \(A\).
  In the case where \(A\) is proven, we are done.
  In the case where \(A\) is disproven, we use lemma~\ref{prop-refute} to
  derive impossibility.
\end{proof}

\begin{lemma} \label{manifest-enum-to-split-enum}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{lemma}
\begin{proof}
  The only difference between manifest enumerability and split enumerability is
  the membership proof: therefor our goal for this proof is to construct a
  function of the following type:
  \begin{equation}
    \lVert x \in \mathit{xs} \rVert \rightarrow x \in \mathit{xs}
  \end{equation}
  Given decidable equality over the type of \(x\).

  We do this using the previous recompute lemma: that tells us that all we need
  to construct is a decision for \(x \in \mathit{xs}\), and it will be able to
  derive the proof itself.
  Such a decision procedure is not difficult to construct: for any value \(x\)
  and list \(\mathit{xs}\), we proceed through the list \(\mathit{xs}\), testing
  if \(x\) is equal to any of its contents.
  If it is, we return that we have proven the goal, and that \(x\) is indeed
  present in \(\mathit{xs}\).
  Otherwise, we know that \(x\) cannot be in \(\mathit{xs}\) (since we've tested
  every value), so we return that the goal has been disproven.
\end{proof}
\section{Kuratowski Finiteness}
The first thing we must define is a representation of subsets.
\begin{definition}[Kuratowski Finite Subset]
  \(\mathcal{K}(A)\) is the type of Kuratowski-finite subsets of \(A\).
  \begin{inductivetype}{\mathcal{K}(A)}
    \inductivetypeclause{\left[ \right]}{\mathcal{K}(A)}
    \inductivetypeclause{\wc \dblcolon \wc}{A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A)}
    \inductivetypeclause{\text{com}}{\Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{dup}}{\Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{trunc}}{\Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q}
  \end{inductivetype}
  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.
